---
title: "Automated Policy Brief: Unemployment Trends"
author: "Nandini Subramanya Hoblidar"
format: html
output-dir: docs
---

# **Policy Brief Generator**

### Introduction

Policy briefs help translate complex data into actionable insights for policymakers, nonprofits, and the public. Traditionally, creating these briefs requires manual data cleaning, analysis, visualization, and writing.

### Why This Project Matters

Policy briefs help translate data into insights for decision-makers. But writing them takes time, and technical findings are often hard to communicate to non-technical audiences.

This project shows how to use **R + Quarto + Large Language Models (LLMs)** to create automated, reproducible policy briefs. The LLM helps generate clear, plain-English summaries of the data analysis, making research more accessible to policymakers, nonprofits, and the public.

### What This Project Does

-   Loads a real dataset (e.g., unemployment rates).\
-   Summarizes and visualizes trends in R.\
-   Calls an LLM to automatically draft a **Policy Implications** section.\
-   Compiles everything into a professional, reproducible report with Quarto.

### How It Can Help

-   **Faculty** can use it as a teaching module on reproducible research and AI-powered reporting.\
-   **Students** can learn both data science workflows and applied AI for communication.

```{r}
#install packages
#install.packages(c("tidyverse", "ggplot2", "readr", "httr", "jsonlite","ellmer","remotes"))

```

Load the necessary packages (tidyverse for data manipulation, ellmer for LLM integration)

Read your CSV file into R

head shows the first 6 rows in your report so readers can see the data

```{r}
# Load libraries
library(tidyverse)
library(ellmer)

# Load unemployment dataset
unemp <- read_csv("data/unemployment.csv", show_col_types = FALSE)
```

```{r}
#col type
spec(unemp)
```

```{r}

# Preview the first few rows
head(unemp)
```

Summary gives a quick overview (min, max, mean, median) of the unemployment trends.

```{r}
# Summary statistics of unemployment rates
summary(unemp$UnemploymentRate)
```

Plots unemployment over time

Shows trends and highlights spikes (like in 2020 for the pandemic)

```{r}
ggplot(unemp, aes(x = Year, y = UnemploymentRate)) +
  geom_line(color = "steelblue", linewidth = 1.5) +
  geom_point(color = "red", size = 3) +
  theme_minimal() +
  labs(title = "Unemployment Rate (2010–2024)",
       x = "Year",
       y = "Unemployment Rate (%)")

```

We want to give the LLM a summary of the dataset in plain text

```{r}
# Create a summary string for the LLM
data_summary <- paste(
  "Unemployment data over the years:\n",
  paste(unemp$Year, unemp$UnemploymentRate, sep=": ", collapse="\n")
)

data_summary
```

The LLM will use the table converted to a string to generate the policy summary

Calling GPT via Ellmer

Creating a prompt to specify the LLM what to do with what data

```{r}

# Create the policy prompt
prompt <- paste(
  "Based on the following unemployment data, write few paragraphs plain-English policy implications summary for policymakers:\n",
  data_summary
)
```

The model to use for the chat defaults to "gpt-4.1" if we don't specify anything.

```{r}
chat <- chat_openai(
  model = NULL,
  system_prompt = prompt)
```

Now we can see the summary of the data provided and reasoning about the unemployment rates. Note that our data has only 2 columns Unemployment Rate and Year respectively.

```{r}
chat$chat(prompt)
```
